---
title: "Appendix"
author: "Kritika and Aurelia"
date: "June 4, 2026"
output:
  pdf_document:
    toc: false
    latex_engine: xelatex
geometry: margin=1in
header-includes:
  - '\usepackage{float}'
  - '\floatplacement{figure}{H}'
  - '\usepackage{booktabs}'
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 6, fig.height = 4, fig.align = "center")
```


# Packages

```{r}
library(tidyverse)
library(VGAM)      
library(ordinal)    
library(MASS)       
library(effects)    
library(broom)       
library(knitr)
library(kableExtra)
```

---

#  Load Data

```{r}
raw <- read.csv("data.csv", na.strings = c("NA", ""), stringsAsFactors = FALSE)
dim(raw)
```

---

# Data Cleaning

## Select Relevant Variables

```{r }
df <- raw %>%
  dplyr::select(ResponseId, JobSat, AISelect, YearsCodePro, EdLevel,
                Employment, ConvertedCompYearly, RemoteWork, Country)
summary(df)
```

## Response Variable: JobSat -> 3 Ordered Categories


```{r }
table(as.numeric(df$JobSat), useNA = "always")

df <- df %>%
  mutate(
    JobSat_num = as.numeric(JobSat),
    JobSat_cat = case_when(
      JobSat_num <= 5  ~ "Low",
      JobSat_num <= 7  ~ "Moderate",
      JobSat_num <= 10 ~ "High",
      TRUE             ~ NA_character_
    ),
    JobSat_cat = fct_relevel(JobSat_cat, "Low", "Moderate")  # High is last/highest
  )

table(df$JobSat_cat, useNA = "always")
```

## YearsCodePro: Recode Text Entries and Center

```{r}
unique(df$YearsCodePro[grepl("year", df$YearsCodePro, ignore.case = TRUE)])

df <- df %>%
  mutate(
    YearsCodePro = case_when(
      YearsCodePro == "Less than 1 year"   ~ "0.5",
      YearsCodePro == "More than 50 years" ~ "50",
      TRUE ~ YearsCodePro
    ),
    YearsCodePro = as.numeric(YearsCodePro)
  )

summary(df$YearsCodePro)
mean(df$YearsCodePro, na.rm = TRUE)

# Center around mean (19.2 years)
df <- df %>% mutate(CYears = YearsCodePro - mean(YearsCodePro, na.rm = TRUE))
```

## Annual Compensation: Log-Transform and Center

Salary is right-skewed with values up to $2M. Log-transforming  then center.

```{r}
sum(df$ConvertedCompYearly > 500000, na.rm = TRUE)  # 25 above $500k
sum(df$ConvertedCompYearly > 1000000, na.rm = TRUE) # 9 above $1M

df <- df %>% mutate(log_comp = log(ConvertedCompYearly))

summary(df$log_comp)
mean(df$log_comp, na.rm = TRUE)

# Center around mean log-comp
df <- df %>% mutate(Clog_comp = log_comp - mean(log_comp, na.rm = TRUE))
```

## Education Level: Collapsing and Level

```{r}
df <- df %>%
  mutate(
    EdLevel = case_when(
      grepl("Bachelor", EdLevel)  ~ "Bachelor's",
      grepl("Master",   EdLevel)  ~ "Master's",
      grepl("Professional degree|Ph\\.D", EdLevel, ignore.case = TRUE) ~ "Professional/PhD",
      grepl("Some college|Associate|Secondary|Primary|Something",
            EdLevel, ignore.case = TRUE) ~ "No Degree",
      TRUE ~ NA_character_
    ),
    EdLevel = fct_relevel(EdLevel, "No Degree", "Bachelor's", "Master's")
  )

table(df$EdLevel, useNA = "always")
```

## Employment Type: Simplifying

```{r }
df <- df %>%
  mutate(
    Employment = case_when(
      grepl("Employed, full-time", Employment)  ~ "Full-time",
      grepl("Independent contractor|freelancer|self-employed",
            Employment, ignore.case = TRUE)     ~ "Freelance",
      grepl("Student",      Employment, ignore.case = TRUE) ~ "Student",
      grepl("part-time",    Employment, ignore.case = TRUE) ~ "Part-time",
      grepl("Not employed", Employment, ignore.case = TRUE) ~ "Not employed",
      TRUE ~ "Other"
    ),
    Employment = fct_relevel(Employment, "Full-time")
  )

table(df$Employment, useNA = "always")
```

## AI Tool Adoption

```{r}
df <- df %>%
  mutate(
    AISelect = case_when(
      AISelect == "Yes"                     ~ "Uses AI",
      AISelect == "No, but I plan to soon"  ~ "Plans to use",
      AISelect == "No, and I don't plan to" ~ "Does not use",
      TRUE ~ NA_character_
    ),
    AISelect = fct_relevel(AISelect, "Does not use", "Plans to use")
  )

table(df$AISelect, useNA = "always")
```

## Remote Work

```{r}
df <- df %>%
  mutate(
    RemoteWork = case_when(
      RemoteWork == "In-person"                         ~ "In-person",
      RemoteWork == "Hybrid (some remote, some in-person)" ~ "Hybrid",
      RemoteWork == "Remote"                            ~ "Remote",
      TRUE ~ NA_character_
    ),
    RemoteWork = fct_relevel(RemoteWork, "In-person", "Hybrid")
  )

table(df$RemoteWork, useNA = "always")
```

## Remove Invalid Country Entries

```{r}
df <- df %>% filter(!Country %in% c("Nomadic", ""))
```

##  Complete Case Analysis

```{r}
df_model <- df %>%
  dplyr::select(ResponseId, JobSat_cat, AISelect, CYears, EdLevel,
                Employment, Clog_comp, RemoteWork, Country) %>%
  drop_na()

cat("Original rows:     ", nrow(df), "\n")
cat("Complete case rows:", nrow(df_model), "\n")
cat("Rows dropped:      ", nrow(df) - nrow(df_model),
    sprintf("(%.1f%%)\n", (nrow(df) - nrow(df_model)) / nrow(df) * 100))
```

## Final Dataset Summary

```{r}
summary(df_model)
```

---

# Exploratory Data Analysis

## Response: Distribution of Job Satisfaction

```{r}
table(df_model$JobSat_cat)
```

```{r}
df_model %>%
  count(JobSat_cat) %>%
  mutate(pct = n / sum(n) * 100) %>%
  ggplot(aes(x = JobSat_cat, y = pct, fill = JobSat_cat)) +
  geom_col(width = 0.6) +
  geom_text(aes(label = sprintf("%.1f%%", pct)), vjust = -0.4, size = 3.5) +
  scale_fill_manual(values = c("Low" = "#d73027", "Moderate" = "#fee090", "High" = "#4575b4")) +
  labs(x = "Job Satisfaction", y = "Percent (%)") +
  theme_bw() + theme(legend.position = "none")
```

## Cumulative Log Odds EDA: Categorical Predictors

### AI Tool Adoption

```{r}
ai_lodds <- df_model %>%
  group_by(AISelect) %>%
  summarize(
    prob1   = mean(JobSat_cat == "Low"),
    lodds1  = log(prob1 / (1 - prob1)),
    prob12  = mean(JobSat_cat %in% c("Low", "Moderate")),
    lodds12 = log(prob12 / (1 - prob12))
  )

ai_lodds
```

```{r}
ggplot(ai_lodds, aes(x = as.numeric(AISelect))) +
  geom_line(aes(y = lodds1,  color = "logit(Low)")) +
  geom_line(aes(y = lodds12, color = "logit(Low, Moderate)")) +
  scale_x_continuous(breaks = 1:3, labels = levels(df_model$AISelect)) +
  labs(x = "AI Tool Adoption", y = "Cumulative log odds", color = "Comparison") +
  theme_bw()
```

### Education Level

```{r}
ed_lodds <- df_model %>%
  group_by(EdLevel) %>%
  summarize(
    prob1   = mean(JobSat_cat == "Low"),
    lodds1  = log(prob1 / (1 - prob1)),
    prob12  = mean(JobSat_cat %in% c("Low", "Moderate")),
    lodds12 = log(prob12 / (1 - prob12))
  )

ed_lodds
```

```{r}
ggplot(ed_lodds, aes(x = as.numeric(EdLevel))) +
  geom_line(aes(y = lodds1,  color = "logit(Low)")) +
  geom_line(aes(y = lodds12, color = "logit(Low, Moderate)")) +
  scale_x_continuous(breaks = 1:4, labels = levels(df_model$EdLevel)) +
  labs(x = "Education Level", y = "Cumulative log odds", color = "Comparison") +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 12, hjust = 1))
```

### Remote Work Status

```{r}
remote_lodds <- df_model %>%
  group_by(RemoteWork) %>%
  summarize(
    prob1   = mean(JobSat_cat == "Low"),
    lodds1  = log(prob1 / (1 - prob1)),
    prob12  = mean(JobSat_cat %in% c("Low", "Moderate")),
    lodds12 = log(prob12 / (1 - prob12))
  )

remote_lodds
```

```{r}
ggplot(remote_lodds, aes(x = as.numeric(RemoteWork))) +
  geom_line(aes(y = lodds1,  color = "logit(Low)")) +
  geom_line(aes(y = lodds12, color = "logit(Low, Moderate)")) +
  scale_x_continuous(breaks = 1:3, labels = levels(df_model$RemoteWork)) +
  labs(x = "Work Arrangement", y = "Cumulative log odds", color = "Comparison") +
  theme_bw()
```

## Cumulative Log Odds EDA: Quantitative Predictors

Group respondents into 5 quantiles, compute mean predictor value and cumulative
log odds per group. Check for linearity and parallel trends.

### Years of Professional Experience

```{r}
exp_lodds <- df_model %>%
  mutate(exp_grp = ntile(CYears, 5)) %>%
  group_by(exp_grp) %>%
  summarize(
    mean_exp = mean(CYears),
    prob1    = mean(JobSat_cat == "Low"),
    lodds1   = log(prob1 / (1 - prob1)),
    prob12   = mean(JobSat_cat %in% c("Low", "Moderate")),
    lodds12  = log(prob12 / (1 - prob12))
  )

exp_lodds
```

```{r}
ggplot(exp_lodds, aes(x = mean_exp)) +
  geom_line(aes(y = lodds1,  color = "logit(Low)")) +
  geom_line(aes(y = lodds12, color = "logit(Low, Moderate)")) +
  geom_point(aes(y = lodds1,  color = "logit(Low)")) +
  geom_point(aes(y = lodds12, color = "logit(Low, Moderate)")) +
  labs(x = "Centered Years of Experience", y = "Cumulative log odds", color = "Comparison") +
  theme_bw()
```

### Log Annual Compensation

```{r}
comp_lodds <- df_model %>%
  mutate(comp_grp = ntile(Clog_comp, 5)) %>%
  group_by(comp_grp) %>%
  summarize(
    mean_comp = mean(Clog_comp),
    prob1     = mean(JobSat_cat == "Low"),
    lodds1    = log(prob1 / (1 - prob1)),
    prob12    = mean(JobSat_cat %in% c("Low", "Moderate")),
    lodds12   = log(prob12 / (1 - prob12))
  )

comp_lodds
```

```{r}
ggplot(comp_lodds, aes(x = mean_comp)) +
  geom_line(aes(y = lodds1,  color = "logit(Low)")) +
  geom_line(aes(y = lodds12, color = "logit(Low, Moderate)")) +
  geom_point(aes(y = lodds1,  color = "logit(Low)")) +
  geom_point(aes(y = lodds12, color = "logit(Low, Moderate)")) +
  labs(x = "Centered Log Annual Compensation", y = "Cumulative log odds",
       color = "Comparison") +
  theme_bw()
```

## Country-Level Variation

```{r}
df_model %>%
  group_by(Country) %>%
  summarise(n = n(), pct_high = mean(JobSat_cat == "High") * 100) %>%
  filter(n >= 20) %>%
  slice_max(n, n = 20) %>%
  ggplot(aes(x = reorder(Country, pct_high), y = pct_high)) +
  geom_col(fill = "#4575b4") +
  coord_flip() +
  labs(x = NULL, y = "% High Job Satisfaction") +
  theme_bw()
```

---

# Modeling


##  Null CLMM: Estimating ICC

```{r}
m0 <- clmm(JobSat_cat ~ 1 + (1 | Country), data = df_model, Hess = TRUE)
summary(m0)

var_u <- as.numeric(VarCorr(m0)$Country)
icc   <- var_u / (var_u + pi^2 / 3)
cat(sprintf("ICC: %.3f — %.1f%% of variation is between countries\n", icc, icc * 100))
```

## Main Effects Model (Model 1)

```{r}
m1 <- clmm(JobSat_cat ~ AISelect + CYears + EdLevel +
              Employment + Clog_comp + RemoteWork + (1 | Country),
            data = df_model, Hess = TRUE)
summary(m1)
```

## Test Random Effect

Comparing Model 1 (with country RE) to fixed-effects-only model via LRT.
Boundary LRTß

```{r}
m1_fixed <- clm(JobSat_cat ~ AISelect + CYears + EdLevel +
                  Employment + Clog_comp + RemoteWork,
                data = df_model)

lrt_obs <- as.numeric(2 * (logLik(m1) - logLik(m1_fixed)))
lrt_obs
# boundary-corrected p-value
.5*(1 - pchisq(lrt_obs, 1)) + .5*(1 - pchisq(lrt_obs, 0))
```

## Proportional Odds Check

Testing whether the proportional odds assumption holds for key predictors
by fitting a partial proportional odds model (parallel = FALSE for AISelect).

```{r}
# Fixed effects partial PO model via VGAM for the PO test
m1_vglm <- vglm(JobSat_cat ~ AISelect + CYears + EdLevel +
                  Employment + Clog_comp + RemoteWork,
                data = df_model, family = cumulative(parallel = TRUE))

m1_partialPO <- vglm(JobSat_cat ~ AISelect + CYears + EdLevel +
                       Employment + Clog_comp + RemoteWork,
                     data = df_model,
                     family = cumulative(parallel = FALSE ~ AISelect))

lrtest(m1_partialPO, m1_vglm)  # H0: PO holds for AISelect
data.frame(model = c("m1_vglm", "m1_partialPO"),
           AIC   = c(AIC(m1_vglm), AIC(m1_partialPO)))
```

## Interaction Models

### Model 2: AI × Experience

```{r}
m2 <- clmm(JobSat_cat ~ AISelect * CYears + EdLevel +
              Employment + Clog_comp + RemoteWork + (1 | Country),
            data = df_model, Hess = TRUE)
summary(m2)
anova(m1, m2)  # LRT: does interaction improve fit?
```

### Model 3: AI × Education

```{r}
m3 <- clmm(JobSat_cat ~ AISelect * EdLevel + CYears +
              Employment + Clog_comp + RemoteWork + (1 | Country),
            data = df_model, Hess = TRUE)
summary(m3)
anova(m1, m3)
```

## Model Comparison

```{r}
AIC(m0, m1, m2, m3)
```

## Final Model


```{r}
m_final <- m1   # update to m2 or m3 if interaction is significant

coef(summary(m_final))
confint(m_final)
```

## Effects Plot (via MASS::polr)

```{r}
df_model <- df_model %>%
  mutate(across(where(is.factor), droplevels),
         JobSat_ord = ordered(JobSat_cat, levels = c("Low", "Moderate", "High")))

m_polr <- MASS::polr(JobSat_ord ~ AISelect + CYears + EdLevel +
                       Employment + Clog_comp + RemoteWork,
                     data = df_model, Hess = TRUE)

# Reference row held at means / most common category
ref_row <- data.frame(
  AISelect   = factor("Uses AI",    levels = levels(df_model$AISelect)),
  CYears     = 0,
  EdLevel    = factor("Bachelor's", levels = levels(df_model$EdLevel)),
  Employment = factor("Full-time",  levels = levels(df_model$Employment)),
  Clog_comp  = 0,
  RemoteWork = factor("Hybrid",     levels = levels(df_model$RemoteWork))
)

eff_plot <- function(var, vals) {
  nd <- ref_row[rep(1, length(vals)), ]
  nd[[var]] <- vals
  rownames(nd) <- NULL
  cbind(nd[var],
        as.data.frame(predict(m_polr, newdata = nd, type = "probs"))) %>%
    pivot_longer(-all_of(var), names_to = "Satisfaction", values_to = "Prob") %>%
    mutate(Satisfaction = factor(Satisfaction, c("Low", "Moderate", "High")))
}

# AISelect
eff_plot("AISelect",
         factor(levels(df_model$AISelect), levels = levels(df_model$AISelect))) %>%
  ggplot(aes(x = AISelect, y = Prob, color = Satisfaction, group = Satisfaction)) +
  geom_line() + geom_point(size = 2) +
  labs(x = "AI Tool Adoption", y = "P(Job Satisfaction)") + theme_bw()

# RemoteWork
eff_plot("RemoteWork",
         factor(levels(df_model$RemoteWork), levels = levels(df_model$RemoteWork))) %>%
  ggplot(aes(x = RemoteWork, y = Prob, color = Satisfaction, group = Satisfaction)) +
  geom_line() + geom_point(size = 2) +
  labs(x = "Remote Work Status", y = "P(Job Satisfaction)") + theme_bw()

# CYears
eff_plot("CYears",
         seq(min(df_model$CYears), max(df_model$CYears), length.out = 50)) %>%
  ggplot(aes(x = CYears, y = Prob, color = Satisfaction)) +
  geom_line() +
  labs(x = "Centered Years of Experience", y = "P(Job Satisfaction)") + theme_bw()
```

```{r}
expand.grid(
  AISelect   = factor(levels(df_model$AISelect),   levels = levels(df_model$AISelect)),
  RemoteWork = factor(levels(df_model$RemoteWork), levels = levels(df_model$RemoteWork)),
  CYears = 0,
  EdLevel    = factor("Bachelor's", levels = levels(df_model$EdLevel)),
  Employment = factor("Full-time",  levels = levels(df_model$Employment)),
  Clog_comp  = 0
) %>%
  cbind(as.data.frame(predict(m_polr, newdata = ., type = "probs"))) %>%
  pivot_longer(c("Low", "Moderate", "High"),
               names_to = "Satisfaction", values_to = "Prob") %>%
  mutate(Satisfaction = factor(Satisfaction, c("Low", "Moderate", "High"))) %>%
  ggplot(aes(x = AISelect, y = Prob, fill = Satisfaction)) +
  geom_col(position = "stack") +
  facet_wrap(~ RemoteWork) +
  scale_fill_manual(
    values = c("Low" = "#d73027", "Moderate" = "#fee090", "High" = "#4575b4")) +
  labs(x = "AI Tool Adoption", y = "P(Job Satisfaction)") + theme_bw()
```

## Diagnostics


```{r}
library(sure)
set.seed(123)
surr_resids <- sure::resids(m_polr, nsim = 50)

# Q-Q plot (replaces autoplot — sure's S3 method conflicts with ggplot2 3.x)
qqnorm(surr_resids,
       main = "Normal Q-Q Plot of Surrogate Residuals",
       ylab = "Surrogate Residuals")
qqline(surr_resids, col = "red", lty = 2)
```

```{r}
par(mfrow = c(1, 2))
plot(df_model$CYears, surr_resids,
     xlab = "Centered Years of Experience", ylab = "Surrogate Residuals")
abline(h = 0, col = "red", lty = 2)

plot(df_model$Clog_comp, surr_resids,
     xlab = "Centered Log Compensation", ylab = "Surrogate Residuals")
abline(h = 0, col = "red", lty = 2)
par(mfrow = c(1, 1))
```

---

# Final Model Results Table

```{r}
coef_out <- as.data.frame(coef(summary(m_final)))
ci_out   <- confint(m_final)

results  <- cbind(coef_out, ci_out)
kable(round(results, 3),
      caption = "Table 1: Final model estimates, standard errors, and 95% CIs.",
      col.names = c("Estimate", "Std. Error", "z-value", "p-value", "2.5%", "97.5%")) %>%
  kable_styling(latex_options = c("hold_position", "striped"), font_size = 9)
```
